Composable data visualizations

### 2020-05-17 wip-compost/paper.tex
Interactive vis and implementation

\subsection{Basic primitives and combinators}

@@

\subsection{Inferring scales and projections}

@@

\section{Interactive charts: Domain-specific event handling}
Data visualizations published on the web often also feature interactivity. Standard forms of
interactivity include animations, hover labels or zooming. More interesting custom forms
include the ``You draw It'' visualization introduced by the New York Times. % cite
The chart shows only the first half of the data, such as a timeline, and the reader has to
guess the second half before clicking a button and seeing the actual data. Standard forms of
interactivity are supported by high-level libraries. Google Charts supports panning using
drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
implemented using low-level libraries such as D3, but doing so requires directly handling
JavaScript events.

@@

Compost can be used in conjunction with a virtual-dom library to create interactive data
visualizations based on the Elm architecture. An interactive visualization is described using
a pair of functions -- the \emph{view} function creates a chart based on the current state and
the \emph{update} function specifies how the state changes when an event occurs.

To illustrate building of interactive data visualizations using Compost, we look at one aspect of
the ``You draw It'' visualizations. We want to create a bar chart where the user can use drag \& drop
to move individual bars. The resulting chart can be seen in Figure~\ref{fig:youdraw}.
The first step is to define types representing the state and events that can occur:
%
\begin{equation*}
\begin{array}{lll}
\fkvd{type}~\ident{State} &\narrow{=}& \ident{bool}~*~(\ident{string}~*~\ident{int})~\ident{list}\\[0.5em]
\fkvd{type}~\ident{Event} &\narrow{=}& \ident{Update}~\fkvd{of}~(\ident{string}~*~\ident{int})\\
                          &\;\narrow{|}& \ident{Moving}~\fkvd{of}~\ident{bool}
\end{array}
\end{equation*}

yadda

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =
  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
\hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
\hspace{1.05em}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
\end{array}
\end{equation*}

yadda

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{render}~t~(m, s)~=\\
\quad \kvd{axis}_l~(\kvd{axis}_b~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{400}),\\
\qquad (~\kvd{mouseMove}~(\lambda \,(\kvd{cat}~p,\_)~(\kvd{cont}~v)\rightarrow \ident{trigger}(\ident{Update}(p, v))),\\
\qquad (~\kvd{mouseUp}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{true}))),\\
\qquad (~\kvd{mouseDown}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{false}))),~\kvd{overlay}~[\\
\qquad \qquad \fkvd{for}~\ident{party}, \ident{mps}~\fkvd{in}~s~\rightarrow~\kvd{padding}~0,10,0,10,~\kvd{fill}~(\ident{color}~\ident{party}),\\
\qquad \qquad\quad\quad [\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp}),\\
\qquad \qquad\quad\quad \hspace{0.38em}(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mps}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,])~~])))))\\
\end{array}
\end{equation*}

\newpage

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/youdraw}
  \vspace{0.25em}
  \caption{You draw.}
  \label{fig:youdraw}
\end{figure}

@@

\newpage

>>>>

\subsection{Basic primitives and combinators}
\label{sec:basic-primitives}

@@

\subsection{Inferring scales and projections}
\label{sec:basic-scales}

@@

\section{Interactive charts: Domain-specific event handling}
Data visualizations published on the web often also feature interactivity. Standard forms of
interactivity include animations, hover labels or zooming. More interesting custom forms
include the ``You draw It'' visualization introduced by the New York Times. % cite
The chart shows only the first half of the data, such as a timeline, and the reader has to
guess the second half before clicking a button and seeing the actual data. Standard forms of
interactivity are supported by high-level libraries. For example, Google Charts supports panning
using drag \& drop, zooming to a selected chart range and animations. Custom interactivity can be
implemented using low-level libraries such as D3, but doing so requires directly handling
JavaScript events.

\subsection{Domain-specific events}
Compost can be used in conjunction with a virtual-dom library to create interactive data
visualizations based on the Elm architecture. An interactive visualization is described using
a pair of functions -- the \emph{view} function creates a chart based on the current state and
the \emph{update} function specifies how the state changes when an event occurs.

@@

\subsection{You draw it data visualizations}
To illustrate building of interactive data visualizations using Compost, we look at one aspect of
the ``You draw It'' visualizations. We want to create a bar chart where the user can use drag \& drop
to move individual bars. The resulting chart can be seen in Figure~\ref{fig:youdraw}.
The first step is to define types representing the state and events that can occur:
%
\begin{equation*}
\begin{array}{lll}
\fkvd{type}~\ident{State} &\narrow{=}& \ident{bool}~*~(\ident{string}~*~\ident{int})~\ident{list}\\[0.5em]
\fkvd{type}~\ident{Event} &\narrow{=}& \ident{Update}~\fkvd{of}~(\ident{string}~*~\ident{int})\\
                          &\;\narrow{|}& \ident{Moving}~\fkvd{of}~\ident{bool}
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
The state is a pair of boolean, indicating whether the user is currently dragging, together with
a list of key value pairs, storing the number of seats for each political party. Two types of events
can occur in the visualization. First, the user may start or stop dragging, which is indicated using
$\ident{Moving}(\fkvd{true})$ and $\ident{Moving}(\fkvd{false})$, respectively. Second, the user
may change a value for a party, which is captured by the $\ident{Update}$ event.

The next part of the visualization is the \ident{update} function which takes an old state together
with an event and produces a new state:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =
  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
\hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
\hspace{1.05em}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
The first two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
represent the party (which bar the user is dragging) and the new value (new number of seats).
If the user is currently dragging, we replace the value associated with the party $p$ in the
list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.
The last case handles the \ident{Moving} event, which simply replaces the first component of the
state tuple, i.e.~a flag indicating whether a mouse button is down.

\begin{figure}[t]
  \includegraphics[scale=0.57]{figures/youdraw}
  \vspace{0.25em}
  \caption{Interactive ``You draw it'' data visualization. The user moves cursor to a bar (left),
    pushes a mouse button and drags the bar to the position that they think is the correct one (right).}
  \label{fig:youdraw}
\end{figure}

Finally, the \ident{render} function takes the current state and builds the data visualization
using the Compost domain-specific language. In addition, it also takes a parameter \ident{trigger},
which is a function of type $\ident{Event}\rightarrow\ident{unit}$ that can be used to trigger
events in handlers, registered using primitives such as \kvd{mouseMove}.
To build the bar chart in Figure~\ref{fig:youdraw}, we use the same approach as in
Section~\ref{sec:basic-primitives}. The only addition are the event handlers registered using
\kvd{mouseMove}, \kvd{mouseUp} and \kvd{mouseDown}:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{render}~\ident{trigger}~(\_, \ident{state})~=\\
\quad \kvd{axis}_l~(\kvd{axis}_b~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{400}),\\
\qquad (~\kvd{mouseMove}~(\lambda \,(\kvd{cat}~p,\_)~(\kvd{cont}~v)\rightarrow \ident{trigger}(\ident{Update}(p, v))),\\
\qquad (~\kvd{mouseUp}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{true}))),\\
\qquad (~\kvd{mouseDown}~(\lambda \,\_~\_\rightarrow \ident{trigger}(\ident{Moving}(\fkvd{false}))),~\kvd{overlay}~[\\
\qquad \qquad \fkvd{for}~\ident{party}, \ident{mps}~\fkvd{in}~\ident{state}~\rightarrow~\kvd{padding}~0,10,0,10,~(\kvd{fill}~(\ident{color}~\ident{party}),\\
\qquad \qquad\quad\quad [\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp}),\\
\qquad \qquad\quad\quad \hspace{0.38em}(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mps}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,])~~])))))\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
When the user interacts with the visualization created using Compost, the library
translates the coordinates associated with events from pixels to domain-specific
values. In case of the above bar chart, when the user moves a mouse, the function registed
using \kvd{mouseMove} is given a categorical value $\kvd{cat}~p, r$ as the X coordinate and
a continuous value $\kvd{cont}~v$ as the Y coordinate. It then takes $p$, which is the name
of the party corresponding to the bar and the value $v$ corresponding to the number of seats
and triggers the $\ident{Update}(p, v)$ event to update the state. The handlers for
\kvd{mouseUp} and \kvd{mouseDown} do not use the coordinates. They simply switch the flag
indicating whether the user is currently dragging or not.

The pair of functions, \ident{update} and \ident{render}, together with an initial state is
all that is needed to create an interactive data visualization. Compost calls
\ident{render} each time the state changes and uses virtual-dom to update the chart
displayed in the browser. Although creating an interactive visualization is more work than
creating a static one, the domain-specific nature of Compost is invaluable. We can
simply take the values $p$ and $v$ produced by a mouse event, use those to update the state
and then, again, render an updated chart.

@@

The Compost library is open-source and the full source code, implmented in the functional
language F\#, can be found at \urrl{http://github.com/compostjs}.
As is often the case with functional domain-specific languages, the implementation is easy once
we find the right collection of basic primitives. This largely applies to the Compost library
and so we will not go into the implementation details. However, it is worth giving an
outline of the implmentation structure.

As mentioned in Section~\ref{sec:basic-scales}, the rendering of shapes proceeds in two stages.
First, the library infers the scales of a shape. When doing so, it also annotates some shapes
with additional information that are needed later for rendering. Second, the library projects
the shape onto an available space and produces the chart, represented as an SVG object.

\subsection{Inferring the scales of a shape}

\begin{equation*}
\begin{array}{rclcl}
s &=& \kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s &|& (\ldots)
\end{array}
\end{equation*}

\begin{equation*}
\ident{calculateScales} ~:~ \ident{Shape} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}
\end{equation*}

\begin{equation*}
\begin{array}{l}
\ident{calculateScales}~(\kvd{nest}_x~v_{min},\;v_{max},\;s) = \\
\quad\fkvd{let}~(s_x, s_y),s' = \ident{calculateScales}~s\\
\quad(\ident{union}~(\ident{singleton}~v_{min})~(\ident{singleton}~v_{max}), s_y),~\kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s'\\
\\
\ident{calculateScales}~(\kvd{overlay}~l) = \\
\quad\fkvd{let}~\ident{scales},\;l'~=~\ident{unzip}~(\ident{map}~\ident{calculateScales}~l)\\
\quad\fkvd{let}~s_x,s_y~=~\ident{unzip}~\ident{scales}\\
\quad(\ident{reduce}~\ident{union}~s_x,~\ident{reduce}~\ident{union}~s_y), \kvd{overlay}~l'
\end{array}
\end{equation*}

\subsection{Drawing shapes}

\begin{equation*}
\ident{project}~:~\ident{float}*\ident{float}\rightarrow \ident{Scale}\rightarrow\ident{Value} \rightarrow \ident{float}
\end{equation*}

\begin{equation*}
\ident{drawShape}~:~\ident{Area} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}\rightarrow\ident{Svg}
\end{equation*}

\begin{equation*}
\begin{array}{l}
\ident{drawShape}~(x_1, y_1, x_2, y_2)~(s_x, s_y)~(\kvd{scaledNest}_x~v_{min},\;v_{max},\;{is}_x,\;\ident{shape}) = \\
\quad\fkvd{let}~x_1' = \ident{project}~(x_1, x_2)~s_x~v_{min}\\
\quad\fkvd{let}~x_2' = \ident{project}~(x_1, x_2)~s_x~v_{max}\\
\quad\ident{drawShape}~(x_1', y_1, x_2', y_2)~({is}_x, s_y)~\ident{shape}\\
\\
\ident{drawShape}~a~s~(\kvd{overlay}~l) = \\
\quad\ident{append}~(\ident{map}~(\ident{drawShape}~a~s)~l)

\end{array}
\end{equation*}

a

b

c

d

e

f

g

h

i
\newpage
